Implementation of Full Synchronous Composition of Discrete Event Models Using IEC 61499 Function Blocks

### 2005-01-31 doc/Sync_Execution_with_IEC61499/paper.tex
*** empty log message ***

Shorter product life-cycles increase the demands on the
manufacturing system. Since the product, to be produced, in
the manufacturing plant is changing frequently the
production equipment needs to be modified too. The frequent
changes of the system make it time consuming and expensive
to update the control system application to reflect the
changes. The same problem applies to both system types
mentioned above and it becomes even larger in mission
critical control applications since the strict requirenment
of guaranteeing the correct operation is imposed. Thus,
there is a need to introduce algorithms and tools to make it
simple and as safe as possible to modify the control system.
The supervisory control theory,\cite{rw:con:1989}, is a
promising framework that might be useful for this purpose.

% suprevisory control
Many man-made systems can, at some abstraction level, be
modeled by discrete event models. A useful class of discrete
event models is the deterministic finite automata which can
be used in suprevisory control theory. In the theory these
models are used, together with a formal specification of the
desired behavior, to automatically generate a control
function which makes the system behave within specification.

@@

% paper organisation
Mathematical definitions that are used in the paper are
presented in the next section. After that, the most
important definitions and results of suprevisory control
theory are introduced followed by an exampel showing how
supervisory control theory can be applied in implementation
of control application for a small system. After the example
a method for implementation of the synchronization operator
for any number of synchronized models is presented together
with the argument for the correctness of the implementation.
In Section~\ref{sec:app_SCT} some considerations for
implementation of supervisory control theory in applications
are presented before the conclusion of the paper.

@@

Synchronous composition, \cite{h:com:1985}, is a useful
approach to model the interaction between automata and is
used in this paper to model the interaction between the
system models.\\
\textbf{Definition}: Full Synchronous Composition of two DFAs\\
The full synchronous composition of two DFAs, $A^1$ and
$A^2$, is denoted ${A}^1||{A}^2=\langle
Q,\Sigma,\delta,q_i,Q_m\rangle$ where $Q=Q^1 \times Q^2$,
$\Sigma=\Sigma^1 \cup \Sigma^2$, $q_i=\langle {q^1_{i}},
{q^2_{i}} \rangle$, $Q_m={Q^1_{m}} \times {Q^2_{m}}$ and the
transition function is
\begin{displaymath}
  \begin{array}{l}
    \delta(\langle{q^1,q^2}\rangle,\sigma)=\\
    =\left\{ {\begin{array}{*{10}l}
          {{ \langle \delta ^1 (q^1 ,\sigma ), \delta ^2 (q^2 ,\sigma ) 
              \rangle}} &  \sigma \in \Gamma^1(q^1) \cap \Gamma^2(q^2) \\
          {{ \langle \delta ^1 (q^1 ,\sigma ), q^2 \rangle} } &  \sigma \in 
          \Gamma^1(q^1) \setminus \Sigma^2 \\
          { { \langle q^1, \delta ^2 (q^2 ,\sigma )\rangle}} & \sigma \in 
          \Gamma^2(q^2) \setminus \Sigma^1 \\
          { \rm undefined} & {{\rm otherwise.}}
        \end{array}} \right.
  \end{array}
\end{displaymath} 
where $q^1 \in Q^1$, $q^2 \in Q^2$ and $\sigma \in \Sigma$.
$\Box$ In other words, if an event is in both alphabets and
it is enabled in both automatons, the transitions associated
with it are made simultaneously in both automatons.
Otherwise, only the automaton containing the event in its
alphabet makes the transition. This definition can be
extended for several automata.\\
\textbf{Definition}: Full Synchronous Composition of $n$ DFAs\\
The full synchronous composition of DFAs, $A^1$, $A^2$ ...
$A^n$, is denoted ${A}^1||...||{A}^n=\langle
Q,\Sigma,\delta,q_i,Q_m\rangle$ where
\begin{eqnarray}
  Q & = & Q^1 \times Q^2 \times ... \times Q^n\nonumber\\
  \Sigma & = & \Sigma^1 \cup \Sigma^2 \cup ... \cup \Sigma^n\nonumber\\
  q_i & = & \langle q^1_i,q^2_i,...,q^n_i \rangle\nonumber\\
  Q_m & = & Q^1_m \times Q^2_m \times ... \times Q^n_m\nonumber
\end{eqnarray}
and the transition function is
\begin{multline}\label{eq:delta}
  \delta(\langle{q^1,...,q^n}\rangle,\sigma)=\\
  =
  \begin{cases}
    {\rm undefined} &{\rm if\ } \exists A^k:\sigma\in\Sigma^k\land\sigma\notin\Gamma^k(q^k)\\
    \langle q^{1+},\dots,q^{n+}\rangle & {\rm otherwise}
  \end{cases} 
\end{multline} 
where 
\begin{equation}\label{eq:qi}
  q^{k+}=
  \begin{cases}
    \delta^k(q^k,\sigma) & {\rm if\ }\sigma\in\Gamma^k(q^k)\\
    q^k                  & {\rm if\ }\sigma\notin\Sigma^k
  \end{cases}
\end{equation}
$\Box$\\
What this definiton states is that in a group of automatons
that share one event if there is one automaton that does not
have that event enabled in its current state, than the event
is not allowed to occur in any of the other automatons in
the group. That is, one automaton is able to prevent all the
others from executing an event if it does not enable it at
the same time as all the other automatons. This property of
the definiton is used to construct a supervisor which
prevents events from occurring. The concept of supervisor is
explained further in the next section which introduces the
supervisory control theory.

\section{Supervisory Control Theory}
\label{sec:sct}
Suprevisory control theory provides algorithms for
calcuation of allowed behavior of the system from discrete
event models of possible behavior and desired behavior. The
model of the possible behavior is called the plant, the
model of the desired behavior is called the specification,
and the result of the computation is called the suprevisor.

In the supervisory control theory, the plant is assumed to
generate all events and the purpose of the supervisor is to
prevent the plant from generating events that will violate
the specfication. The assumption that the plant generates
all events might seem like a restriction. However, it is
often possible to introduce a controller which will initiate
the events between the supervisor and the resource. Treating
the controller as part of the plant will make the assumption
that all events are generated by the plant a natural one. In
Fig.~\ref{fig:sup_plant} such configuration is shown.
\begin{figure}
  \centering \includegraphics[scale=0.6]{figures/sup_plant}
  \caption{Supervisory control configuration.}
  \label{fig:sup_plant}
\end{figure}

@@

% IEC 61499 intro
The emerging standard IEC 61499 specifies a generic
architecture for measurement and control of industrial
processes. The architecture is based on functional software
units called function blocks. The most common type of
function block in an application is basic function block and
it is shown in Fig.~\ref{fig:61499_intro} using the
graphical representation. This type of block executes
algorithms based on the arriving events and generates new
events that are passed on when the algorithms finish
execution. The algorithms use data associated with incoming
events to update the internal variables and produce the
output data associated with the outgoing events. Which
algorithms are invoked is controlled by the execution
control chart (ECC) which is an automaton with algorithms
associated with different states. When a state is entered
the algorithms associated with it are executed once and the
ECC stays in the state until a condition for leaving is
fulfilled. The conditions upon which transitions occur are
boolean expressions involving event input variables,
internal variables, input variables, and output variables.
When the expression evaluates to \texttt{true} the
transition occur. Special case is transition labled with
\texttt{1}, it is taken as soon as the algorithms finish
execution.
\begin{figure}
  \centering
  \includegraphics[scale=0.4]{figures/basic_block}
  \caption{Anatomy of a basic function block type.}
  \label{fig:61499_intro}
\end{figure}

@@

The control application connecting all of the function
blocks together is shown in Fig.~\ref{fig:app}. Small
circles in the figure are event multiplexer and
demultiplexer. Event multiplexer generates an event on the
output every time an event arrives at any one of the inputs.
Event demultiplexer does the opossite and generates an event
on all of the outputs whenever an event arrives at the
input.
\begin{figure}
  \centering
  \includegraphics[scale=0.45]{figures/app}
  \caption{The control application for the example.}
  \label{fig:app}
\end{figure}

@@

When algorithm is generated what is left is to connect
instances of the model function blocks with the
synchronization function block to generate the control
application. Some further considerations for the
implementaion of supervisory control theory are presented in
the following section.

@@

Today implementation of a distributed control application is
done using application structure similar to the structure in
Fig.~\ref{fig:old_app_struct}. Supervisor \texttt{S} in
the application is manualy crafted using informal
specificatons of the system behavior, and transfered to the
PLC-code. For each application the communication between the
suprevisor and the plants has to be implemented in a
application specific way since the existing standards for
communication between PLCs are at a low level of
abstraction. Using IEC 61499 the communication part of the
application is abstracted far away from the hardware level
and the control engineer is only considering the event and
data passing between the functional parts of the
application, i.e. function block instances. When the system
plants are upgraded or chaged the procedure of supervisor
crafting has to be done once again and very small parts of
the code from the application already implemented are
reused, i.e. the communication parts.
\begin{figure}
  \centering 
  \includegraphics[scale=0.6]{figures/old_app_struct}
  \caption{The structure of distributed control applications today.}
  \label{fig:old_app_struct}
\end{figure}

In software tool Supremica \cite{aff:sup:2003,www:sup:2004},
algorithms from supervisory control theory are implemented
so that a supervisor can be synthesised from automata models
of the plants and specifications. The algorithm that
implements the general method for synchronization according
to this paper is also implemented in the same tool. What
this means is that the control engineer can focus on the
correct modeling of the plants and transfering of the
informal specifications to the specification models for use
in the Supremica tool. Then the tool automatically generates
the supervisor in form of an automaton model, function block
types that implement the skeleton for the plants, the
supervisor automaton, synchronous composition function block
and finally the control application that connects the
instances of the generated function block types as seen in
Fig.~\ref{fig:app} for the example. The sekelton in this
case refers to the fact that the communication with the
hardware has to be implemented by hand but all the behavior
of the models is automatically generated. Using the tool to
generate the control application gives new application
structure as shown in Fig.~\ref{fig:new_app_struct}. In
this structure, only the models for the generation of the
plant blocks and the hardware control portion of those
blocks have to be upgraded whenever the plants in the system
change. All other parts of the application are regenerated
automatically.
\begin{figure}[b]
  \centering 
  \includegraphics[scale=0.6]{figures/new_app_struct}
  \caption{A new structure for distributed control applications.}
  \label{fig:new_app_struct}
\end{figure}

When trying to implement the control application for complex
systems with many plants and specifications the problem of
state space explosion is encountered when trying to
synthesise the supervisor model. There are methods that
represent the plants, specifications and the supervisor in
different ways to try to cope with the problem. For example,
a supervisor may be expressed as a singel automaton,
multiple automata, and a binary decision diagram for
off-line synthesis. Or the supervisor may be synthesized
on-line while the control application is executing. The
method presented in this paper can be used for
synchronization of all types of representations since the
model interface (see Fig.~\ref{fig:model_interface}) is
based on events which are a general characteristic of the
discrete event models.

\section{Conclusion}
Using supervisory control theory in construction of control
applications can significantly reduce development time but
until now it was dificult to apply the theory to
implementation of distributed control applications. Using
the emerging standard IEC 61499 it becomes significantly
easier. The method presented in this paper brings
supervisory control theory together with distributed control
application development in an efficient way.

>>>>

Shorter product life-cycles increase the demands on the
manufacturing system. Since the product, to be produced, in
the manufacturing plant is changing frequently the
production equipment needs to be modified too. The frequent
changes of the system make it time consuming and expensive
to update the control system application to reflect the
changes. The same problem applies to both system types
mentioned above and it becomes even larger in mission
critical control applications since the strict requirement
of guaranteeing the correct operation is imposed. Thus,
there is a need to introduce algorithms and tools to make it
simple and as safe as possible to modify the control system.
The supervisory control theory,\cite{rw:con:1989}, is a
promising framework that might be useful for this purpose.

% supervisory control
Many man-made systems can, at some abstraction level, be
modeled by discrete event models. A useful class of discrete
event models is the deterministic finite automata which can
be used in supervisory control theory. In the theory these
models are used, together with a formal specification of the
desired behavior, to automatically generate a control
function which makes the system behave within specification.

@@

% paper organization
Mathematical definitions that are used in the paper are
presented in the next section. After that, the most
important definitions and results of supervisory control
theory are introduced followed by an example showing how
supervisory control theory can be applied in implementation
of control application for a small system. After the example
a method for implementation of the synchronization operator
for any number of synchronized models is presented together
with the argument for the correctness of the implementation.
In Section~\ref{sec:app_SCT} some considerations for
implementation of supervisory control theory in applications
are presented before the conclusion of the paper.

@@

Synchronous composition, \cite{h:com:1985}, is a useful
approach to model the interaction between automata and is
used in this paper to model the interaction between the
system models.\\
\textbf{Definition}: Full Synchronous Composition of two DFAs\\
The full synchronous composition of two DFAs, $A^1$ and
$A^2$, is denoted ${A}^1||{A}^2=\langle
Q,\Sigma,\delta,q_i,Q_m\rangle$ where $Q=Q^1 \times Q^2$,
$\Sigma=\Sigma^1 \cup \Sigma^2$, $q_i=\langle {q^1_{i}},
{q^2_{i}} \rangle$, $Q_m={Q^1_{m}} \times {Q^2_{m}}$ and the
transition function is
\begin{displaymath}
  \begin{array}{l}
    \delta(\langle{q^1,q^2}\rangle,\sigma)=\\
    =\left\{ {\begin{array}{*{10}l}
          {{ \langle \delta ^1 (q^1 ,\sigma ), \delta ^2 (q^2 ,\sigma ) 
              \rangle}} &  \sigma \in \Gamma^1(q^1) \cap \Gamma^2(q^2) \\
          {{ \langle \delta ^1 (q^1 ,\sigma ), q^2 \rangle} } &  \sigma \in 
          \Gamma^1(q^1) \setminus \Sigma^2 \\
          { { \langle q^1, \delta ^2 (q^2 ,\sigma )\rangle}} & \sigma \in 
          \Gamma^2(q^2) \setminus \Sigma^1 \\
          { \rm undefined} & {{\rm otherwise.}}
        \end{array}} \right.
  \end{array}
\end{displaymath} 
where $q^1 \in Q^1$, $q^2 \in Q^2$ and $\sigma \in \Sigma$.
$\Box$ In other words, if an event is in both alphabets and
it is enabled in both automatons, the transitions associated
with it are made simultaneously in both automatons.
Otherwise, only the automaton containing the event in its
alphabet makes the transition. This definition can be
extended for several automata.\\
\textbf{Definition}: Full Synchronous Composition of $n$ DFAs\\
The full synchronous composition of DFAs, $A^1$, $A^2$ ...
$A^n$, is denoted ${A}^1||...||{A}^n=\langle
Q,\Sigma,\delta,q_i,Q_m\rangle$ where
\begin{eqnarray}
  Q & = & Q^1 \times Q^2 \times ... \times Q^n\nonumber\\
  \Sigma & = & \Sigma^1 \cup \Sigma^2 \cup ... \cup \Sigma^n\nonumber\\
  q_i & = & \langle q^1_i,q^2_i,...,q^n_i \rangle\nonumber\\
  Q_m & = & Q^1_m \times Q^2_m \times ... \times Q^n_m\nonumber
\end{eqnarray}
and the transition function is
\begin{multline}\label{eq:delta}
  \delta(\langle{q^1,...,q^n}\rangle,\sigma)=\\
  =
  \begin{cases}
    {\rm undefined} &{\rm if\ } \exists A^k:\sigma\in\Sigma^k\land\sigma\notin\Gamma^k(q^k)\\
    \langle q^{1+},\dots,q^{n+}\rangle & {\rm otherwise}
  \end{cases} 
\end{multline} 
where 
\begin{equation}\label{eq:qi}
  q^{k+}=
  \begin{cases}
    \delta^k(q^k,\sigma) & {\rm if\ }\sigma\in\Gamma^k(q^k)\\
    q^k                  & {\rm if\ }\sigma\notin\Sigma^k
  \end{cases}
\end{equation}
$\Box$\\
What this definition states is that in a group of automatons
that share one event if there is one automaton that does not
have that event enabled in its current state, than the event
is not allowed to occur in any of the other automatons in
the group. That is, one automaton is able to prevent all the
others from executing an event if it does not enable it at
the same time as all the other automatons. This property of
the definition is used to construct a supervisor which
prevents events from occurring. The concept of supervisor is
explained further in the next section which introduces the
supervisory control theory.

\section{Supervisory Control Theory}
\label{sec:sct}
Supervisory control theory provides algorithms for
calculation of allowed behavior of the system from discrete
event models of possible behavior and desired behavior. The
model of the possible behavior is called the plant, the
model of the desired behavior is called the specification,
and the result of the computation is called the supervisor.

In the supervisory control theory, the plant is assumed to
generate all events and the purpose of the supervisor is to
prevent the plant from generating events that will violate
the specification. The assumption that the plant generates
all events might seem like a restriction. However, it is
often possible to introduce a controller which will initiate
the events between the supervisor and the resource. Treating
the controller as part of the plant will make the assumption
that all events are generated by the plant a natural one. In
Fig.~\ref{fig:sup_plant} such configuration is shown.
\begin{figure}
  \centering \includegraphics[scale=0.6]{figures/sup_plant}
  \caption{Supervisory control configuration.}
  \label{fig:sup_plant}
\end{figure}

@@

% IEC 61499 intro
The emerging standard IEC 61499 specifies a generic
architecture for measurement and control of industrial
processes. The architecture is based on functional software
units called function blocks. The most common type of
function block in an application is basic function block and
it is shown in Fig.~\ref{fig:61499_intro} using the
graphical representation. This type of block executes
algorithms based on the arriving events and generates new
events that are passed on when the algorithms finish
execution. The algorithms use data associated with incoming
events to update the internal variables and produce the
output data associated with the outgoing events. Which
algorithms are invoked is controlled by the execution
control chart (ECC) which is an automaton with algorithms
associated with different states. When a state is entered
the algorithms associated with it are executed once and the
ECC stays in the state until a condition for leaving is
fulfilled. The conditions upon which transitions occur are
boolean expressions involving event input variables,
internal variables, input variables, and output variables.
When the expression evaluates to \texttt{true} the
transition occur. Special case is transition labeled with
\texttt{1}, it is taken as soon as the algorithms finish
execution.
\begin{figure}
  \centering
  \includegraphics[scale=0.4]{figures/basic_block}
  \caption{Anatomy of a basic function block type.}
  \label{fig:61499_intro}
\end{figure}

@@

The control application connecting all of the function
blocks together is shown in Fig.~\ref{fig:app}. Small
circles in the figure are event multiplexer and
demultiplexer. Event multiplexer generates an event on the
output every time an event arrives at any one of the inputs.
Event demultiplexer does the opposite and generates an event
on all of the outputs whenever an event arrives at the
input.
\begin{figure}
  \centering
  \includegraphics[scale=0.45]{figures/app}
  \caption{The control application for the example.}
  \label{fig:app}
\end{figure}

@@

When algorithm is generated what is left is to connect
instances of the model function blocks with the
synchronization function block to generate the control
application. Some further considerations for the
implementation of supervisory control theory are presented in
the following section.

@@

Today implementation of a distributed control application is
done using application structure similar to the structure in
Fig.~\ref{fig:old_app_struct}. Supervisor \texttt{S} in
the application is manually crafted using informal
specifications of the system behavior, and transfered to the
PLC-code. For each application the communication between the
supervisor and the plants has to be implemented in a
application specific way since the existing standards for
communication between PLCs are at a low level of
abstraction. Using IEC 61499 the communication part of the
application is abstracted far away from the hardware level
and the control engineer is only considering the event and
data passing between the functional parts of the
application, i.e. function block instances. When the system
plants are upgraded or changed the procedure of supervisor
crafting has to be done once again and very small parts of
the code from the application already implemented are
reused, i.e. the communication parts.
\begin{figure}
  \centering 
  \includegraphics[scale=0.6]{figures/old_app_struct}
  \caption{The structure of distributed control applications today.}
  \label{fig:old_app_struct}
\end{figure}

In software tool Supremica \cite{aff:sup:2003,www:sup:2004},
algorithms from supervisory control theory are implemented
so that a supervisor can be synthesized from automata models
of the plants and specifications. The algorithm that
implements the general method for synchronization according
to this paper is also implemented in the same tool. What
this means is that the control engineer can focus on the
correct modeling of the plants and transferring of the
informal specifications to the specification models for use
in the Supremica tool. Then the tool automatically generates
the supervisor in form of an automaton model, function block
types that implement the skeleton for the plants, the
supervisor automaton, synchronous composition function block
and finally the control application that connects the
instances of the generated function block types as seen in
Fig.~\ref{fig:app} for the example. The skeleton in this
case refers to the fact that the communication with the
hardware has to be implemented by hand but all the behavior
of the models is automatically generated. Using the tool to
generate the control application gives new application
structure as shown in Fig.~\ref{fig:new_app_struct}. In
this structure, only the models for the generation of the
plant blocks and the hardware control portion of those
blocks have to be upgraded whenever the plants in the system
change. All other parts of the application are regenerated
automatically.
\begin{figure}[b]
  \centering 
  \includegraphics[scale=0.6]{figures/new_app_struct}
  \caption{A new structure for distributed control applications.}
  \label{fig:new_app_struct}
\end{figure}

When trying to implement the control application for complex
systems with many plants and specifications the problem of
state space explosion is encountered when trying to
synthesize the supervisor model. There are methods that
represent the plants, specifications and the supervisor in
different ways to try to cope with the problem. For example,
a supervisor may be expressed as a single automaton,
multiple automata, and a binary decision diagram for
off-line synthesis. Or the supervisor may be synthesized
on-line while the control application is executing. The
method presented in this paper can be used for
synchronization of all types of representations since the
model interface (see Fig.~\ref{fig:model_interface}) is
based on events which are a general characteristic of the
discrete event models.

\section{Conclusion}
Using supervisory control theory in construction of control
applications can significantly reduce development time but
until now it was difficult to apply the theory to
implementation of distributed control applications. Using
the emerging standard IEC 61499 it becomes significantly
easier. The method presented in this paper brings
supervisory control theory together with distributed control
application development in an efficient way.
